Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-09-02 TACO-paper-v2/coffee.tex
minor fix
small rewriting of generality section

We first notice that, in the scenario $[f=1, p=1]$, the assembly is dominated by matrix insertion: despite the application of several transformations, only a minimal performance gain is achieved. This changes instantly moving to $p=2$ and $f=2$. In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible. In such cases, the transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}. Interesting is the fact that generalized loop-invariant code motion was particularly invasive, with 23 temporaries generated and several redundancies discovered (see Section~\ref{sec:code-transf-licm}).

@@

\section{Generality of the Approach and Applicability to Other Domains}
\label{sec:generality}
We have demonstrated that our optimizations for cross-loop arithmetic intensity are effective in the context of automated code generation for finite element local assembly. In this section, we discuss about the potential for their applicability in other computational domains and, more in general, to what extent a general-purpose compiler could integrate some of them.

We observe that our choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2. This was motivated by two critical aspects.

@@

\paragraph{Generalization to other domains} There are neither conceptual nor technical reasons for which our compiler could not be used outside of the Firedrake system. For example, integration with the popular FEniCS framework, the pioneer of automated code generation for finite element local assembly, would be relatively easy to achieve. More challenging is to assess the generality of our optimization strategy; that is, if and how COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent this would be helpful for improving full-application performance. To answer these questions, we first need to go back to the origins of our compiler. The starting point of our work was the mathematical formulation of a local assembly operation, expressible as follows

@@

The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$ to compute the local element matrix $A$. Functions $\alpha$, $\beta$ and $\gamma$ are problem-specific and can be intricately complex, involving for example the evaluation of derivatives. Abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted

\paragraph{Optimizing mathematical expressions} 
Expression manipulation at the mathematical layer (e.g. simplification, decomposition into sub-expressions) opens multiple semantically-equivalent code generation opportunities, characterized by different trade-offs in parallelism, redundant computation, and data locality. The basic idea is to exploit properties of arithmetic operators, like associativity and commutativity, to re-schedule the computation suitably for the underlying architecture. Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domain computationally similar to local assembly, like those involving tensor contractions~\cite{TCE}. We highlight three notable points.

@@

\item In Equation~\ref{eq:assembly-model}, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. Throughout the paper we assumed to have a single thread per kernel, although it is worth noting that for certain architectures (e.g. GPUs) this could be prohibitive due to memory requirements. Intra-kernel parallelization is one possible solution: a domain-specific compiler like COFFEE could map mathematical quantifiers and operators to different parallelization schemes, and transform both the kernel and the function calling it for multi-threaded execution. This approach is fundamental if performance portability needs to be achieved. We believe that analogous considerations apply to different computational domains.

@@

\paragraph{Code generation and applicability of the transformations} 
All array sizes and loop bounds, like $n1$ and $n2$ in Equation~\ref{eq:assembly-model}, are known at code generation time. This means that ``good'' code can be generated. For example, loop bounds can be made explicit, arrays can be statically initialized, and pointer aliasing is easily avoidable. Further, all of these factors contribute to the applicability and the effectiveness of all of our code transformations. For instance, knowing loop bounds allows both generating correct code when applying vector-register tiling (see Section~\ref{sec:op-vect}) and discovering redundant computation opportunities (see Section~\ref{sec:code-transf-licm}). On the other hand, the padding and data alignment transformation could be performed directly at run-time; note that, theoretically, it could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.  
\paragraph{Multi-loop vectorization} 
Compiler auto-vectorization has become effective for a variety of codes. Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers to the best of our knowledge. However, the outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization; more importantly, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi results). Despite being not straightforward to automate multi-loop vectorization in a general-purpose compiler, especially if stencils are present, we believe that in specific domains this could be more easily achieved. The intuition is to map the memory access pattern on vector registers, and then relying on in-register shuffling to minimize the traffic between memory and processor.

>>>>

We first notice that, in the scenario $[f=1, p=1]$, the assembly is dominated by matrix insertion: despite the application of several transformations, only a minimal performance gain is achieved. This changes instantly increasing $p$ or $f$. In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible. In such cases, the transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}. Interesting is the fact that generalized loop-invariant code motion was particularly invasive, with 23 temporaries generated and several redundancies discovered (see Section~\ref{sec:code-transf-licm}).

@@

\section{Generality of the Approach and Applicability to Other Domains}
\label{sec:generality}
We have demonstrated that our cross-loop optimizations for arithmetic intensity are effective in the context of automated code generation for finite element local assembly. In this section, we discuss about their applicability in other computational domains and, more in general, their integrability within a general-purpose compiler.

We observe that our choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2. This was motivated by two critical aspects that characterize the generality of our research.

@@

\paragraph{Generalizability to other domains} There are neither conceptual nor technical reasons for which our compiler could not be used outside of Firedrake. For example, integration with the popular FEniCS framework, the pioneer of automated code generation for finite element local assembly, would be relatively easy to achieve. More challenging is to assess the generality of the optimization strategy; that is, if and how COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent this would be helpful for improving full-application performance. To answer these questions, we first need to go back to the origins of our compiler. The starting point of our work was the mathematical formulation of a local assembly operation, expressible as follows

@@

The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$ computing the local element matrix $A$. Functions $\alpha$, $\beta$ and $\gamma$ are problem-specific and can be intricately complex, involving for example the evaluation of derivatives. We can however abstract from the inherent structure of $\alpha$, $\beta$ and $\gamma$ to highlight a number of aspects

\begin{itemize}
\item \textbf{Optimizing mathematical expressions.} Expression manipulation (e.g. simplification, decomposition into sub-expressions) opens multiple semantically-equivalent code generation opportunities, characterized by different trade-offs in parallelism, redundant computation, and data locality. The basic idea is to exploit properties of arithmetic operators, like associativity and commutativity, to re-schedule the computation suitably for the underlying architecture. Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domains involving numerical evaluation of complex mathematical expressions (e.g. electronic structure calculations in physics and quantum chemistry relying on tensor contractions~\cite{TCE}). In this context, we highlight three notable points.

@@

\item In Equation~\ref{eq:assembly-model}, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. Throughout the paper we assumed to assign the execution of a kernel to a single thread, which is likely to be the best strategy for a standard multi-core CPU. On the other hand, we note that for certain architectures (e.g. GPUs) this could be prohibitive due to memory requirements. Intra-kernel parallelization is one possible solution: a domain-specific compiler like COFFEE could map mathematical quantifiers and operators to different parallelization schemes and generate distinct variants of multi-threaded kernel code. Based on our experience, we believe this is the right approach to achieve performance portability. 

@@

\item \textbf{Code generation and applicability of the transformations.} All array sizes and loop bounds, like $n1$ and $n2$ in Equation~\ref{eq:assembly-model}, are known at code generation time. This means that ``good'' code can be generated. For example, loop bounds can be made explicit, arrays can be statically initialized, and pointer aliasing is easily avoidable. Further, all of these factors contribute to the applicability and the effectiveness of all of our code transformations. For instance, knowing loop bounds allows both generating correct code when applying vector-register tiling (see Section~\ref{sec:op-vect}) and discovering redundant computation opportunities (see Section~\ref{sec:code-transf-licm}). On the other hand, the padding and data alignment transformation could be performed directly at run-time; note that, theoretically, it could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.  
\item \textbf{Multi-loop vectorization.} Compiler auto-vectorization has become effective for a variety of codes. Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers to the best of our knowledge. However, the outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization; more importantly, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi results). Despite being not straightforward to automate multi-loop vectorization in a general-purpose compiler, especially if stencils are present, we believe that in specific domains this could be more easily achieved. The intuition is to map the memory access pattern on vector registers, and then relying on in-register shuffling to minimize the traffic between memory and processor.
\end{itemize}
